Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-03-25 Manuscript/detrendify_aoas.tex
Re-organized ADMM section, added discussion of detrendr_cross

In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, the three sharp transient spikes at 11:32, 14:10, and 16:03, but the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering
and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis.

@@

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold.
\begin{itemize}
	\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure valid trend estimates.
	\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. The algorithm proceeds by splitting the time domain into overlapping windows, fitting the model separately for each of the windows and reconciling estimates from the overlapping intervals.
	\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

@@

%We review \textit{$\ell_1$-trend filtering} for estimating a mean trend before discussing how to adapt it to estimate a baseline trend. We then describe a new scalable algorithm for simultaneously computing multiple quantile trends over large times windows.

@@

%	$k+1$ fold composition of the discrete difference operator produces estimates that are piecewise polynomials of order $k$.

%\cite{Tib2014} later demonstrated empirically that the trend filtering estimates adapt to the local level of smoothness better than smoothing splines and also proved that with a judicious choice of tuning parameter the trend filtering estimate converges to the true underlying function at the minimax rate for functions whose $k$th derivative is of bounded variation.
%
%	Given observations $y(t)$ with $t=1,...,N$ and defining $y \in \Real^n$ such that $y_i = y(t)$, in the trend filtering problem, the trend, $\V{\theta} \in \Real^n$, is estimated by solving the following convex problem:
%	\begin{eqnarray}
%	\underset{\V{\theta}}{\arg\min}\; \frac{1}{2} \lVert y - \V{\theta} \rVert_2^2 + \lambda \lVert \Mn{D}{k+1}\V{\theta} \rVert_1,
%	\end{eqnarray}
%	where $y \in \Real^n$ is the vector of observations, $\lambda \geq 0$ is a regularization parameter, and the matrix $\Mn{D}{k+1} \in \Real^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\Mn{D}{k+1}$ consider the difference operator when $k = 0$.
%	\begin{eqnarray*}
%	\Mn{D}{1} & = & \begin{pmatrix}
%
%	-1 & 1 & 0 & \cdots & 0 & 0 \\
%	0 & -1 & 1 & \cdots & 0 & 0 \\
%	\vdots & & & & & \\
%	0 & 0 & 0 & \cdots & -1 & 1 \\
%
%	\end{pmatrix}.
%	\end{eqnarray*}
%	Thus, $\lVert \Mn{D}{1}\V{\theta} \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_i - \theta_{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty term incentivizes solutions which are piecewise constant. For $k \geq 1$, the difference operator $\Mn{D}{k+1} \in \Real^{(n-k-1) \times n}$ is defined recursively as follows
%	\begin{eqnarray*}
%	\Mn{D}{k+1} & = & \Mn{D}{1}\Mn{D}{k}.
%	\end{eqnarray*}
%	By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piecewise polynomials of order $k$.

\cite{Tib2014} proved that with a judicious choice of $\lambda$ the trend filtering estimate converges to the true underlying function at the minimax rate for functions whose $k$th derivative is of bounded variation and showed that trend filtering is both fast and locally adaptive when the time series consists of only the trend and random noise, which is illustrated in \Fig{tpn}. As noted earlier though in some applications, such as the air quality monitoring problem considered in this paper, the data contain a rapidly varying signal in addition to the slowly varying trend and noise. \Fig{tpspn} shows that standard trend filtering is not designed to distinguish between the slowly varying trend and the rapidly-varying signal, as the smooth component estimate $\V{\theta}$ is biased towards the peak of the transient component.

To account for the presence of a transient component in the observed time series $\V{y}$, we propose quantile trend filtering \Fig{tpspn}. % as a more robust method for detrending in these situations \Fig{tpspn}.
To estimate the trend in the $\tau$\textsuperscript{th} quantile, we solve the convex optimization problem
\begin{eqnarray}
\label{eq:quantile_trend}
\underset{\V{\theta}}{\text{minimize}}\; \rho_\tau(\V{y} - \V{\theta}) + \lambda \lVert \Mn{D}{k+1} \V{\theta} \rVert_1,
\end{eqnarray}
where $\rho_\tau(\V{r})$ is the check function
\begin{eqnarray}
\label{eq:check}
\rho_{\tau}(\V{r}) & = & \sum_{i=1}^n \VE{r}{i}(1-\One(\VE{r}{i}<0)),
\end{eqnarray}
and $\One(A)$ is 1 if its input $A$ is true and 0 otherwise.

\begin{figure}
	\centering
	\subfloat[Slow Trend + Noise]{\label{fig:tpn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}}
	\subfloat[Slow Trend + Spikes + Noise]{\label{fig:tpspn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}}
	\caption{Examples of trend filtering solutions (red) and 15\textsuperscript{th} quantile trending filtering solution (blue). Standard trend filtering performs well in the no-signal case (right) but struggles to distinguish between the slowly varying trend and the rapidly-varying signal (left). The quantile trend is not affected by the signal and provides an estimate of the baseline.}
\end{figure}
Note that we do not explicitly model $\V{s}$. Rather, we focus on estimating $\V{\theta}$. We then estimate $\V{s} + \V{\varepsilon}$ as the difference $\V{y} - \V{\theta}$.

Before elaborating on how we compute our proposed $\ell_1$-quantile trend filtering estimator, we discuss similarities and differences between our proposed estimator and existing baseline estimators.

@@

\begin{equation*}
\label{eq:smoothingspline}
\sum_{i=1}^n\rho_{\tau}(y(t_i) - \theta(t_i)) + \lambda\int \theta''(t)^2 dt,
\end{equation*}
where $\theta(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check function with a differentiable function. \cite{Racine2017} take a different approach and constrain the response to follow a location scale model and estimate the $\tau$\textsuperscript{th} conditional quantile using a kernel estimator and local linear approach. \Halley{Eric}{What are $a(t), b(t)$, and $\epsilon_i$?}

@@

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray*}
	\label{eq:noncross_trend}
	%	&&\underset{\V{\theta}_1, \ldots, \V{\theta}_J}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	%	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
	%	&&\text{subject to: }\; \V{\theta}_1 \leq \V{\theta}_2 \leq \ldots \leq \V{\theta}_J
	%	\theta_{1}(t) \le \theta_{2}(t) \le \ldots \le \theta_{J}(t) \text{ for all } t, \nonumber
\end{eqnarray*}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, eventually the matrix of constraints will no longer fit into memory. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem.

\begin{equation}
\label{eq:split_objective}
\begin{split}
&\text{minimize} \; f(\V{\theta}) + g(\Vtilde{\theta}) \\
&\text{subject to} \; \M{A}\V{\theta} + \M{B}\Vtilde{\theta} = \V{c}.
\end{split}
\end{equation}
\Eric{Joe}{I'm already lost here. Can we give some context to what $\theta$ and $\widetilde{\theta}$ are in the context of our windowing algorithm? Maybe we should say something about the windowing first before we dive head first into Lagrangians? What are $A$ and $B$? Do these simplify to something in our specific application? The stuff below might seem obvious to you, but I'm totally lost. It might be better to talk about the notation for the windows first, and how different windows will give different thetas for the overlapping parts, and then talk about equality constrained optimization as a solution for the overlapping parts.} Recall that finding the minimizer to an equality constrained optimization problem is equivalent to the identifying the saddle point of the Lagrangian function associated with the problem \Eqn{split_objective}. ADMM seeks the saddle point of a related function called the augmented Lagrangian,
%The Lagrangian for the ALM problem
\begin{eqnarray*}
	\mathcal{L}_{\gamma}(\V{\theta},\Vtilde{\theta},\V{\omega}) & = & f(\V{\theta}) + g(\Vtilde{\theta}) + \langle \V{\omega}, \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rangle
	+ \frac{\gamma}{2} \lVert \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rVert_2^2,
\end{eqnarray*}
where the dual variable $\V{\omega}$ is a vector of Lagrange multipliers and $\gamma$ is a nonnegative tuning parameter. When $\gamma = 0$, the augmented Lagrangian coincides with the ordinary Lagrangian.

@@

Let $\mathcal{I}$ denote a subset of the indices $\{1, \ldots, n\}$ and $\V{y}_\mathcal{I} \in \Real^{\lvert I \rvert}$ denote the subvector obtained by stacking on top of each other the elements $\VE{y}{i}$ for $i \in \mathcal{I}$ in their canonical ordering.
We apply the consensus ADMM algorithm to the quantile regression trend filtering problem given in \Eqn{quantile_trend}, by dividing our observed series $\VE{y}{i}$ with $i = 1, \ldots, n$ into overlapping $W$ windows of observations $\Vn{y}{w} = \V{y}_{\mathcal{I}_w}$ where $\mathcal{I}_w$ is the set of integers from $l_w$ to $u_w$ inclusive, namely $\mathcal{I}_w = \{l_w, l_w + 1, \ldots, u_w -1, u_w\}$ where
\begin{eqnarray*}
	1 = l_{1} < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < \cdots < u_{W} = n.
\end{eqnarray*}
\Fig{windows} shows an example of 1200 observations being mapped into three equally sized overlapping windows of observations.

%	\begin{align*}
%	y_1(t) = y(t) & \mbox{~~if~~} 1 \le t \le u_{1}\\
%	y_2(t) = y(t) & \mbox{~~if~~} l_{2} \le t \le u_{2} \\
%	\vdots & \\
%	y_M(t) = y(t) & \mbox{~~if~~} l_{M} \le t \le  n,
%	\end{align*}
%
%	 with boundaries $1 < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < ...< n$.
%	 Given quantiles $\tau_1 < \cdots < \tau_J$, we define $\theta_{j,m}(t)$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $m$ at time $t$. In order to constrain the overlapping sections to be equal, we define a consensus variable
Given quantiles $\tau_1 < \cdots < \tau_J$, we introduce dummy variables $\Vn{\theta}{w}_{j} \in \Real^{\lvert \mathcal{I}_w \rvert}$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $w$. We then ``stitch" together the $W$ quantile trend estimates into consensus over the overlapping regions by introducing the constraint $\VnE{\theta}{w}{ij} = \VE{\theta}{i + l_w - 1,j}$ for $i = 1, \ldots, u_w - l_w + 1$ and for all $j$. Let $\Mn{\Theta}{w}$ be the matrix whose $j$th column is $\Vn{\theta}{w}$. Then we can write these constraints more compactly as $\Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta}$, where $\Mn{U}{w} \in \{0, 1\}^{\lvert \mathcal{I}_w \rvert \times n}$ is a matrix that selects rows of $\M{\Theta}$ corresponding to the $w$th window, namely
\begin{eqnarray*}
	\Mn{U}{w} & = & \begin{pmatrix}
		\V{e}_{l_w}\Tra \\
		\vdots \\
		\V{e}_{u_w}\Tra
	\end{pmatrix},
\end{eqnarray*}
where $\V{e}_i \in \Real^{n}$ denotes the $i$th standard basis vector. Furthermore, let $\iota_{\mathcal{C}}$ denote the indicator function of the non-crossing quantile constraint, namely $\iota_{\mathcal{C}}(\M{\Theta})$ is zero if $\M{\Theta} \in \mathcal{C}$ and infinity otherwise.

Our windowed quantile trend optimization problem can then be written as
\begin{equation}
\label{eq:quantile_windows}
\begin{split}
&\text{minimize}\; \sum_{w=1}^W \left \{\sum_{j=1}^J \left [\rho_{\tau_j}(\Vn{y}{w} - \Vn{\theta}{w}_{j}) +
\lambda \lVert \Mn{D}{k+1} \Vn{\theta}{w}_{j} \rVert_1 \right ] + \iota_\mathcal{C}(\Mn{\Theta}{w}) \right\}\\
&\text{subject to} \qquad \Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta} \;\; \text{ for } w = 1, \ldots, W.
\end{split}
\end{equation}

Let $\Mn{\Omega}{w}$ denote the Lagrange multiplier matrix for the $w$th consensus constraint, namely $\Mn{\theta}{w} = \Mn{U}{w}\M{\theta}$, and let $\Vn{\omega}{w}_{j}$ denote its $j$th column.
%
%		  $\V{\omega}_{m}_{j}$ denote Lagrange multipliers for the constraint $\Vn{\theta}{w}_j = \Mn{U}{m}\V{\theta}_j$, $\Mn{\Omega}{m}$ denote the matrix whose $j$th column is $\V{\omega}_{m}_j$, and penalty parameter $\gamma > 0$,

The augmented Lagrangian is given by
\begin{eqnarray*}
	\mathcal{L}(\M{\Theta}, \{\Mn{\Theta}{w}\}, \{\Mn{\Omega}{w}\}) & = & \sum_{w=1}^W \mathcal{L}_w(\M{\Theta}, \Mn{\Theta}{w}, \Mn{\Omega}{w}),
\end{eqnarray*}
where
\begin{eqnarray*}
	\mathcal{L}_w(\M{\Theta}, \Mn{\Theta}{w}, \Mn{\Omega}{w}) & = & \sum_{j=1}^J \biggl [\rho_{\tau_j}(\Vn{y}{w} - \Vn{\theta}{w}_j)+\lambda \lVert \Mn{D}{k+1}\Vn{\theta}{w}_j\rVert_1 \\
	&& + \; (\Vn{\theta}{w}_j - \Mn{U}{w}\V{\theta}_{j})\Tra\Vn{\omega}{w}_{j} +
	\frac{\gamma}{2}\lVert \Vn{\theta}{w}_j - \Mn{U}{w}\V{\theta}_{j}\rVert_2^2 \biggr ] + \iota_\mathcal{C}(\Mn{\Theta}{w}),
\end{eqnarray*}
where $\gamma$ is a positive penalty parameter.

The ADMM algorithm alternates between updating the consensus variable $\M{\Theta}$, the window variables $\{\Mn{\Theta}{w}\}$, and the Lagrange multipliers $\{\Mn{\Omega}{w}\}$.
At the $m+1$th iteration, we perform the following sequence of updates

@@

\subsection{Validation}
Our method can easily handle missing data by defining the check loss function to output 0 for missing values. %We also explore three information criteria.
Specifically, we use the following modified function $\tilde{\rho}_\tau$ in place of the $\rho_\tau$ function given in \Eqn{check}
\begin{eqnarray}
\label{eq:modcheck}
\tilde{\rho}_{\tau}(\V{r}) & = & \sum_{i \not\in \mathcal{V}} \VE{r}{i}(1-\One(\VE{r}{i}<0)),
\end{eqnarray}
where $\mathcal{V}$ is a held-out validation subset of $\{1, \ldots, n\}$ and solve the problem

>>>>

In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) was not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering
and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis.

@@

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold.
\begin{itemize}
	\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure validity and improve trend estimates.
	\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. The algorithm proceeds by splitting the time domain into overlapping windows, fitting the model separately for each of the windows and reconciling estimates from the overlapping intervals.
	\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

@@

@@

\cite{Tib2014} proved that with a judicious choice of $\lambda$ the trend filtering estimate converges to the true underlying function at the minimax rate for functions whose $k$th derivative is of bounded variation and showed that trend filtering is both fast and locally adaptive when the time series consists of only the trend and random noise, which is illustrated in \Fig{tpn}. As noted earlier, in some applications, such as the air quality monitoring problem considered in this paper, the data contain a rapidly varying signal in addition to the slowly varying trend and noise. \Fig{tpspn} shows that standard trend filtering is not designed to distinguish between the slowly varying trend and the rapidly-varying signal, as the smooth component estimate $\V{\theta}$ is biased towards the peak of the transient component.

To account for the presence of a transient component in the observed time series $\V{y}$, we propose quantile trend filtering \Fig{tpspn}. % as a more robust method for detrending in these situations \Fig{tpspn}.
To estimate the trend in the $\tau$\textsuperscript{th} quantile, we solve the convex optimization problem
\begin{eqnarray}
\label{eq:quantile_trend}
\underset{\V{\theta}}{\text{minimize}}\; \rho_\tau(\V{y} - \V{\theta}) + \lambda \lVert \Mn{D}{k+1} \V{\theta} \rVert_1,
\end{eqnarray}
where $\rho_\tau(\V{r})$ is the check function
\begin{eqnarray}
\label{eq:check}
\rho_{\tau}(\V{r}) & = & \sum_{i=1}^n \VE{r}{i}(\tau-\One(\VE{r}{i}<0)),
\end{eqnarray}
and $\One(A)$ is 1 if its input $A$ is true and 0 otherwise. Note that we do not explicitly model $\V{s}$. Rather, we focus on estimating $\V{\theta}$. We then estimate $\V{s} + \V{\varepsilon}$ as the difference $\V{y} - \V{\theta}$.

\begin{figure}
	\centering
	\subfloat[Slow Trend + Noise]{\label{fig:tpn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}}
	\subfloat[Slow Trend + Spikes + Noise]{\label{fig:tpspn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}}
	\caption{Examples of trend filtering solutions (red) and 15\textsuperscript{th} quantile trending filtering solution (blue). Standard trend filtering performs well in the no-signal case (a) but struggles to distinguish between the slowly varying trend and the rapidly-varying signal (b). The quantile trend is not affected by the signal and provides an estimate of the baseline.}
\end{figure}

Before elaborating on how we compute our proposed $\ell_1$-quantile trend filtering estimator, we discuss similarities and differences between our proposed estimator and existing quantile trend estimators.

@@

\begin{equation*}
\label{eq:smoothingspline}
\sum_{i=1}^n\rho_{\tau}(y(t_i) - \theta(t_i)) + \lambda\int \theta''(t)^2 dt,
\end{equation*}
where $\theta(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check function with a differentiable function. \cite{Racine2017} take a different approach and constrain the response to follow a location scale model and estimate the $\tau$\textsuperscript{th} conditional quantile by combining Gaussian quantile functions with a kernel smoother and solving a local-linear least squares problem.

@@

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimated multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray*}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
\end{eqnarray*}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. We allow for the possibilty that the degree of smoothness in the trends varies by quantile and thus allow the smoothing parameter as well. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

 As the size of the data increases, all methods require additional computation time and for our method eventually the matrix of constraints no longer fits into memory. In our application to air quality sensor data, measurements are recorded every second resulting in 86,400 observations per day. This number of observations is already too large to use with currently available R packages used for estimating quantile trends \citep{fields, quantreg}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion.

 \begin{figure}[!t]
 	\centering
 	\includegraphics[width = 0.8\linewidth]{Figures/overlapping_windows.png}
 	\caption{Window boundaries and trends fit separately in each window.}
 	\label{fig:windows}
 \end{figure}

 Let $\mathcal{I}$ denote a subset of the indices $\{1, \ldots, n\}$ and $\V{y}_\mathcal{I} \in \Real^{\lvert I \rvert}$ denote the subvector obtained by stacking on top of each other the elements $\VE{y}{i}$ for $i \in \mathcal{I}$ in their canonical ordering.
 To decrease computation and extend our method to larger problems, we divide our observed series $\VE{y}{i}$ with $i = 1, \ldots, n$ into $W$ overlapping windows of observations $\Vn{y}{w} = \V{y}_{\mathcal{I}_w}$ where $\mathcal{I}_w$ is the set of integers from $l_w$ to $u_w$ inclusive, namely $\mathcal{I}_w = \{l_w, l_w + 1, \ldots, u_w -1, u_w\}$ and
 \begin{eqnarray*}
 	1 = l_{1} < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < \cdots < u_{W} = n.
 \end{eqnarray*}
 \Fig{windows} shows an example of 1200 observations being mapped into three equally sized overlapping windows of observations. While the overlapping trend estimates between $l_2$ and $u_2$ don't vary dramatically, the difference is more pronounced in the trend in the 5\textsuperscript{th} quantile between $l_3$ and $u_2$.

 Given quantiles $\tau_1 < \cdots < \tau_J$, we introduce dummy variables $\Vn{\theta}{w}_{j} \in \Real^{\lvert \mathcal{I}_w \rvert}$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $w$. We then ``stitch" together the $W$ quantile trend estimates into consensus over the overlapping regions by introducing the constraint $\VnE{\theta}{w}{ij} = \VE{\theta}{i + l_w - 1,j}$ for $i = 1, \ldots, u_w - l_w + 1$ and for all $j$. Let $\Mn{\Theta}{w}$ be the matrix whose $j$th column is $\Vn{\theta}{w}_{j}$. Then we can write these constraints more compactly as $\Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta}$, where $\Mn{U}{w} \in \{0, 1\}^{\lvert \mathcal{I}_w \rvert \times n}$ is a matrix that selects rows of $\M{\Theta}$ corresponding to the $w$th window, namely
 \begin{eqnarray*}
 	\Mn{U}{w} & = & \begin{pmatrix}
 		\V{e}_{l_w}\Tra \\
 		\vdots \\
 		\V{e}_{u_w}\Tra
 	\end{pmatrix},
 \end{eqnarray*}
 where $\V{e}_i \in \Real^{n}$ denotes the $i$th standard basis vector. Furthermore, let $\iota_{\mathcal{C}}$ denote the indicator function of the non-crossing quantile constraint, namely $\iota_{\mathcal{C}}(\M{\Theta})$ is zero if $\M{\Theta} \in \mathcal{C}$ and infinity otherwise.

 Our windowed quantile trend optimization problem can then be written as
 \begin{equation}
 \label{eq:quantile_windows}
 \begin{split}
 &\text{minimize}\; \sum_{w=1}^W \left \{\sum_{j=1}^J \left [\rho_{\tau_j}(\Vn{y}{w} - \Vn{\theta}{w}_{j}) +
 \lambda_j \lVert \Mn{D}{k+1} \Vn{\theta}{w}_{j} \rVert_1 \right ] + \iota_\mathcal{C}(\Mn{\Theta}{w}) \right\}\\
 &\text{subject to} \qquad \Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta} \;\; \text{ for } w = 1, \ldots, W.
 \end{split}
 \end{equation}

 The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem which is a more general form of \Eqn{quantile_windows}.

\begin{equation}
\label{eq:split_objective}
\begin{split}
&\text{minimize} \; f(\V{\theta}) + g(\Vtilde{\theta}) \\
&\text{subject to} \; \M{A}\V{\theta} + \M{B}\Vtilde{\theta} = \V{c}.
\end{split}
\end{equation}
Recall that finding the minimizer to an equality constrained optimization problem is equivalent to the identifying the saddle point of the Lagrangian function associated with the problem \Eqn{split_objective}. ADMM seeks the saddle point of a related function called the augmented Lagrangian,
%The Lagrangian for the ALM problem
\begin{eqnarray*}
	\mathcal{L}_{\gamma}(\V{\theta},\Vtilde{\theta},\V{\omega}) & = & f(\V{\theta}) + g(\Vtilde{\theta}) + \langle \V{\omega}, \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rangle
	+ \frac{\gamma}{2} \lVert \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rVert_2^2,
\end{eqnarray*}
where the dual variable $\V{\omega}$ is a vector of Lagrange multipliers and $\gamma$ is a nonnegative tuning parameter. When $\gamma = 0$, the augmented Lagrangian coincides with the ordinary Lagrangian.

@@

Returning to our constrained windows problem giving in \Eqn{quantile_windows}, let $\Mn{\Omega}{w}$ denote the Lagrange multiplier matrix for the $w$th consensus constraint, namely $\Mn{\theta}{w} = \Mn{U}{w}\M{\theta}$, and let $\Vn{\omega}{w}_{j}$ denote its $j$th column.

The augmented Lagrangian is given by
\begin{eqnarray*}
	\mathcal{L}(\M{\Theta}, \{\Mn{\Theta}{w}\}, \{\Mn{\Omega}{w}\}) & = & \sum_{w=1}^W \mathcal{L}_w(\M{\Theta}, \Mn{\Theta}{w}, \Mn{\Omega}{w}),
\end{eqnarray*}
where
\begin{eqnarray*}
	\mathcal{L}_w(\M{\Theta}, \Mn{\Theta}{w}, \Mn{\Omega}{w}) & = & \sum_{j=1}^J \biggl [\rho_{\tau_j}(\Vn{y}{w} - \Vn{\theta}{w}_j)+\lambda_j \lVert \Mn{D}{k+1}\Vn{\theta}{w}_j\rVert_1 \\
	&& + \; (\Vn{\theta}{w}_j - \Mn{U}{w}\V{\theta}_{j})\Tra\Vn{\omega}{w}_{j} +
	\frac{\gamma}{2}\lVert \Vn{\theta}{w}_j - \Mn{U}{w}\V{\theta}_{j}\rVert_2^2 \biggr ] + \iota_\mathcal{C}(\Mn{\Theta}{w}),
\end{eqnarray*}
where $\gamma$ is a positive penalty parameter.

The ADMM algorithm alternates between updating the consensus variable $\M{\Theta}$, the window variables $\{\Mn{\Theta}{w}\}$, and the Lagrange multipliers $\{\Mn{\Omega}{w}\}$.
At the $(m+1)$\textsuperscript{th} iteration, we perform the following sequence of updates

@@

\subsection{Validation}
Our method can easily handle missing data by defining the check loss function to output 0 for missing values. %We also explore three information criteria.
Specifically, we use the following modified function $\tilde{\rho}_\tau$ in place of the $\rho_\tau$ function given in \Eqn{check}
\begin{eqnarray}
\label{eq:modcheck}
\tilde{\rho}_{\tau}(\V{r}) & = & \sum_{i \not\in \mathcal{V}} \VE{r}{i}(\tau-\One(\VE{r}{i}<0)),
\end{eqnarray}
where $\mathcal{V}$ is a held-out validation subset of $\{1, \ldots, n\}$ and solve the problem
